% ============================================================================
\chapter{Introduction}
% ============================================================================

\section{Philosophy}

Caffeine Engine is built under a single principle: \emph{transparency}.
Every byte that passes through the CPU must be accountable to the
programmer. This philosophy manifests in three concrete decisions.

\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Zero-dependency standard library.} The engine replaces
        \texttt{std} containers with its own implementations tuned for
        game-loop access patterns (cache-friendly sequential iteration,
        allocator-aware growth, no hidden heap traffic).

  \item \textbf{Data-oriented design.} Entities are not objects; they are
        integer identifiers. Components are plain-old-data structs stored in
        contiguous typed pools. This layout maximises SIMD and cache line
        utilisation during system updates.

  \item \textbf{Explicit concurrency.} No implicit background threads.
        The job system is the single scheduling primitive; callers declare
        dependencies explicitly through barriers and handles.
\end{itemize}

\section{Document Structure}

Each subsequent chapter covers one major subsystem. The ordering follows
the dependency graph: fundamental types and mathematics are presented first,
followed by memory, containers, ECS, threading, domain systems (physics,
audio), the asset pipeline, and finally the editor viewport. Each chapter
concludes with a table of invariants that any future implementation must
preserve.

\section{Notation Conventions}

Throughout this document, the following notation is used consistently.

\begin{itemize}[leftmargin=1.5em]
  \item Scalars: italic Latin letters --- $a, b, t, \theta$.
  \item Vectors: bold lower-case --- $\mathbf{v}, \mathbf{u}, \mathbf{n}$.
  \item Matrices: bold upper-case --- $\mathbf{M}, \mathbf{R}$.
  \item Quaternions: calligraphic --- $\mathcal{q}$.
  \item Unit vectors: hat notation --- $\hat{\mathbf{v}}$.
  \item Column-major storage is assumed unless stated otherwise.
  \item Code identifiers appear in \texttt{monospace}.
\end{itemize}
