% ============================================================================
\chapter{Spatial Audio System}
% ============================================================================

\section{Architecture}

The audio system wraps SDL3's \texttt{SDL\_AudioStream} API. It maintains
a pool of \texttt{AudioSource} objects (default 32 SFX channels).
Each source maps to one \texttt{SDL\_AudioStream} bound to the system's
\texttt{SDL\_AudioDeviceID} (default device, S16 stereo 44100\,Hz).

\section{AudioClip}

An \texttt{AudioClip} is a pure descriptor: a pointer to raw PCM bytes,
size, sample rate, channel count, bits per sample, and duration.
The system does not own the memory; clips are registered by the asset
pipeline which manages their lifetimes.

\section{Spatial Attenuation}

For an emitter at world position $\mathbf{p}_e$ and listener at
$\mathbf{p}_l$ with maximum audible distance $d_\text{max}$:

\begin{equation}
\text{volume} = \max\!\left(0,\; 1 - \frac{|\mathbf{p}_e - \mathbf{p}_l|}{d_\text{max}}\right)
\label{eq:audio-volume}
\end{equation}

This is a \emph{linear falloff} model. Physically accurate inverse-square
falloff ($1/d^2$) sounds too aggressive for game contexts; linear falloff
produces a smoother perceptual transition.

\section{Stereo Panning}

The pan value in $[-1, 1]$ is derived from the horizontal offset of the
emitter relative to the listener:

\begin{equation}
\text{pan} = \text{clamp}\!\left(\frac{p_{ex} - p_{lx}}{d_\text{max}},\; -1,\; 1\right)
\end{equation}

Gain for each channel:
\begin{align}
  g_L &= v \cdot (1 - \max(0, \text{pan})) \\
  g_R &= v \cdot (1 + \min(0, \text{pan}))
\end{align}

The stream gain is set to $\max(g_L, g_R)$. True per-channel gain
requires a custom mixing callback (planned).

\section{Playback Loop}

When a looping source exhausts its buffer (SDL reports
\texttt{SDL\_GetAudioStreamAvailable} == 0), the full clip data is
re-submitted via \texttt{SDL\_PutAudioStreamData}. The implementation
tracks elapsed time to detect this condition frame-accurately.
