% ============================================================================
\chapter{Asset Pipeline (.caf Format)}
% ============================================================================

\section{Design Goals}

The Caffeine Asset Format (\texttt{.caf}) is a \textbf{zero-parsing,
zero-copy} binary container. The layout on disk is a direct mirror of
the in-RAM / in-VRAM layout, so loading is a single \texttt{fread} followed
by pointer arithmetic --- no deserialization step, no allocations beyond
the buffer itself.

\section{File Layout}

\begin{equation*}
\underbrace{[0\ldots 31]}_{\text{CafHeader}}
\underbrace{[32\ldots 32+N-1]}_{\text{Metadata}}
\underbrace{[32+N\ldots 32+N+M-1]}_{\text{Payload}}
\underbrace{[32+N+M\ldots 32+N+M+3]}_{\text{Footer CRC32}}
\end{equation*}

\begin{table}[H]
\centering
\caption{CafHeader fields (32 bytes, little-endian)}
\begin{tabular}{lllr}
\toprule
\textbf{Offset} & \textbf{Field} & \textbf{Type} & \textbf{Bytes} \\
\midrule
0  & \texttt{magic}        & \texttt{u32} (= 0xCAFECAFE) & 4 \\
4  & \texttt{versionMajor} & \texttt{u16}                & 2 \\
6  & \texttt{versionMinor} & \texttt{u16}                & 2 \\
8  & \texttt{type}         & \texttt{AssetType} (u16)    & 2 \\
10 & \texttt{flags}        & \texttt{CafFlags} (u16)     & 2 \\
12 & \texttt{crc32}        & \texttt{u32}                & 4 \\
16 & \texttt{metadataSize} & \texttt{u64}                & 8 \\
24 & \texttt{dataSize}     & \texttt{u64}                & 8 \\
\bottomrule
\end{tabular}
\end{table}

\begin{specbox}{Invariant 10.1 --- Endianness}
  The \texttt{.caf} format is little-endian.
  A \texttt{static\_assert} verifies \texttt{std::endian::native == std::endian::little}
  at compile time; the format must not be used on big-endian platforms
  without byte-swapping shims.
\end{specbox}

\section{Integrity Verification}

Two CRC32 checks are performed at load time:
\begin{enumerate}
  \item \textbf{Payload CRC32}: \texttt{crc32(payload, dataSize)} must equal
        \texttt{header.crc32}.
  \item \textbf{Whole-file CRC32}: The 4-byte footer stores
        \texttt{crc32(everything-before-footer)}, providing end-to-end
        integrity against file corruption or truncation.
\end{enumerate}

The CRC32 uses the IEEE 802.3 polynomial $P = \texttt{0xEDB88320}$
(reflected form). The lookup table of 256 entries is built at
compile time via a \texttt{constexpr} function, adding zero run-time
initialisation cost.

\begin{specbox}{Invariant 10.2 --- CRC32 verification}
  Both CRC32 checks must pass before any \texttt{.caf} file is used.
  A file that passes only one check is treated as corrupt.
\end{specbox}

\section{Asset Types}

\begin{table}[H]
\centering
\caption{AssetType discriminants and their metadata structs}
\begin{tabular}{lll}
\toprule
\textbf{Type} & \textbf{Metadata struct} & \textbf{Notes} \\
\midrule
\texttt{Texture}   & \texttt{TextureMetadata} (24 B)  & width, height, format, mips \\
\texttt{Audio}     & \texttt{AudioMetadata} (16 B)    & sampleRate, channels, samples \\
\texttt{Mesh}      & (Phase 5)                        & vertex / index buffers \\
\texttt{Prefab}    & \texttt{SceneMetadata} (20 B)    & binary ECS entity template \\
\texttt{Scene}     & \texttt{SceneMetadata} (20 B)    & full world state snapshot \\
\texttt{Shader}    & \texttt{ShaderMetadata} (8 B)    & stage (vert/frag/compute) \\
\texttt{Animation} & (planned)                        & animation clip frames \\
\bottomrule
\end{tabular}
\end{table}

\section{Live Asset Watching}

\texttt{FileWatcher} monitors a directory via
\texttt{ReadDirectoryChangesW} (Windows) or a polling stub (other
platforms). Changed paths are pushed onto a thread-safe queue and polled
from the main thread so that the asset loader can hot-reload without
cross-thread world mutation.
