% ============================================================================
\chapter{Implementation Rules and Future Specifications}
% ============================================================================

\section{Binding Invariants}

The following invariants must be preserved by all future implementations.
A proposed change that violates any invariant requires explicit discussion
and this document must be updated before implementation begins.

\begin{table}[H]
\centering
\caption{Binding invariants summary}
\begin{tabular}{lp{9cm}}
\toprule
\textbf{ID} & \textbf{Statement} \\
\midrule
I-1  & All type widths validated by \texttt{static\_assert}. \\
I-2  & \texttt{CF\_ASSERT} is never removed to silence errors. \\
I-3  & Allocator alignment is always a power of two $\geq 8$. \\
I-4  & Pool slot size is never exceeded at the call site. \\
I-5  & Component IDs are registered atomically; no component may receive two different IDs. \\
I-6  & ECS structural mutations during iteration go through \texttt{CommandBuffer}. \\
I-7  & Job handles must check version equality before reporting completion. \\
I-8  & \texttt{.caf} files are always verified with both CRC32 checks before use. \\
I-9  & Near-plane clipping is applied to every projected line segment in Mode3D. \\
I-10 & Navigation widget axes are computed from live \texttt{camYaw}/\texttt{camPitch} values, never hardcoded. \\
\bottomrule
\end{tabular}
\end{table}

\section{Planned Systems}

The following systems are specified here to constrain future
implementation choices, even though they are not yet implemented.

\subsection{Mesh Rendering (Phase 5)}

Mesh assets store interleaved vertex data (position, normal, UV, tangent)
aligned to 32 bytes for AVX2 SIMD. The RHI layer (\texttt{src/rhi/})
wraps SDL3-GPU command buffers; mesh rendering will use indexed draws
with a single persistent vertex buffer per mesh, uploaded once at asset
load time.

\subsection{Scripting (Lua 5.4)}

The scripting engine (\texttt{src/script/}) embeds Lua 5.4.
Each entity may have a \texttt{ScriptComponent} that stores a Lua table
as its instance state. The engine exposes the ECS world to scripts via a
C API binding layer. Script execution runs on the main thread; no
cross-thread Lua state sharing is permitted.

\subsection{Animation}

Animation clips store keyframe data (time, value, tangent for Hermite
interpolation) per channel (position, rotation, scale). The
\texttt{AnimationSystem} evaluates clips via the Hermite spline formula
and writes results into transform components.
