% ============================================================================
\chapter*{Bibliography}
\addcontentsline{toc}{chapter}{Bibliography}
% ============================================================================

\begin{thebibliography}{99}

\bibitem{sdl_gpu}
SDL3 GPU Category, official documentation.
\url{https://wiki.libsdl.org/SDL3/CategoryGPU}

\bibitem{jylanki}
J. Jylänki, ``A Thousand Ways to Pack the Bin, A Practical Approach to Two-Dimensional Rectangle Bin Packing,'' 2010.

\bibitem{fiedler}
G. Fiedler, ``Fix Your Timestep!,'' \emph{Gaffer on Games}, 2004.
\url{https://gafferongames.com/post/fix_your_timestep/}

\bibitem{shoemake}
K. Shoemake, ``Animating Rotation with Quaternion Curves,'' \emph{SIGGRAPH 1985 Proceedings}, ACM, pp. 245, 254, 1985.

\bibitem{chase_lev}
D. Chase and Y. Lev, ``Dynamic Circular Work-Stealing Deque,'' \emph{SPAA 2005}, pp. 21, 28, 2005.

\bibitem{baumgarte}
J. Baumgarte, ``Stabilization of Constraints and Integrals of Motion in Dynamical Systems,'' \emph{Computer Methods in Applied Mechanics and Engineering}, vol. 1, no. 1, pp. 1, 16, 1972.

\bibitem{crc32}
CRC-32 IEEE 802.3 Standard (Ethernet), CRC-32 Polynomial definition. Also referenced in \emph{IETF RFC 3720}.

\bibitem{opengl}
Khronos Group, ``OpenGL Specification,'' column-major matrix convention.
\url{https://www.khronos.org/opengl/}

\bibitem{eberly}
D. Eberly, \emph{3D Game Engine Architecture}, Morgan Kaufmann, 2005.

\bibitem{mirtich}
B. Mirtich, ``Impulse-Based Dynamic Simulation of Rigid Body Systems,'' Ph.D. thesis, University of California, Berkeley, 1996.

\bibitem{gems6}
M. Dickheiser (ed.), \emph{Game Programming Gems 6}, Charles River Media, 2006.

\bibitem{imgui}
Dear ImGui, \url{https://github.com/ocornut/imgui}

\bibitem{cpp20}
ISO/IEC 14882:2020, \emph{Programming Languages, C++}, 2020.

\end{thebibliography}
